---
title: "Supplementary Figure 2 - RFhap computational resource profile"
author: "RFhap manuscript"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dev = "png",
  dpi = 150
)

required_packages <- c(
  "dplyr", "tidyr", "readr", "stringr", "forcats",
  "lubridate", "ggplot2", "scales", "patchwork", "cowplot", "ragg"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Missing R packages: ", paste(missing_packages, collapse = ", "),
    ". Install them before rendering this document."
  )
}

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(stringr)
  library(forcats)
  library(lubridate)
  library(ggplot2)
  library(scales)
  library(patchwork)
})

input_file <- file.path("data", "execution_trace_HG002.txt")
output_dir <- "outputs"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

if (!file.exists(input_file)) {
  stop("Required input file was not found: ", input_file)
}
```

## Objective

This document reconstructs Supplementary Figure 2 directly from the HG002
Nextflow execution trace. It summarizes CPU time, memory, task-level wall time,
and disk I/O for the RFhap processes contributing at least 0.5 CPU-hours.

## Parsing functions

```{r parsing-functions}
parse_duration_seconds <- function(x) {
  x <- replace_na(as.character(x), "0s")
  vapply(x, function(value) {
    hits <- str_match_all(
      str_trim(value),
      "([0-9.]+)\\s*(ms|d|h|m|s)"
    )[[1]]

    if (nrow(hits) == 0) {
      return(suppressWarnings(as.numeric(value)))
    }

    multiplier <- c(ms = 0.001, d = 86400, h = 3600, m = 60, s = 1)
    sum(as.numeric(hits[, 2]) * unname(multiplier[hits[, 3]]), na.rm = TRUE)
  }, numeric(1))
}

parse_bytes <- function(x) {
  x <- replace_na(as.character(x), "0")
  x <- str_trim(x)
  x[x == "0"] <- "0 B"

  hit <- str_match(x, "^([0-9.]+)\\s*([KMGTPE]?B)$")
  value <- as.numeric(hit[, 2])
  unit <- str_to_upper(hit[, 3])

  multiplier <- c(
    B = 1,
    KB = 1024,
    MB = 1024^2,
    GB = 1024^3,
    TB = 1024^4,
    PB = 1024^5,
    EB = 1024^6
  )

  value * unname(multiplier[unit])
}

clean_task_name <- function(x) {
  str_trim(str_replace(x, "\\s*\\([0-9]+\\)$", ""))
}
```

## Load and process the execution trace

```{r load-and-process}
trace <- read_tsv(
  input_file,
  show_col_types = FALSE,
  progress = FALSE,
  name_repair = "minimal"
)

task_metrics <- trace %>%
  filter(status %in% c("COMPLETED", "CACHED")) %>%
  mutate(
    process = clean_task_name(name),
    process = recode(process, create_kmers_database = "kmc3"),
    submit_time = ymd_hms(submit, tz = "UTC"),
    realtime_seconds = parse_duration_seconds(realtime),
    duration_seconds = parse_duration_seconds(duration),
    cpu_percent = as.numeric(str_remove(`%cpu`, "%")),
    cpu_time_seconds = realtime_seconds * cpu_percent / 100,
    peak_rss_GiB = parse_bytes(peak_rss) / 1024^3,
    peak_vmem_GiB = parse_bytes(peak_vmem) / 1024^3,
    read_GiB = parse_bytes(rchar) / 1024^3,
    write_GiB = parse_bytes(wchar) / 1024^3,
    cpu_hours = cpu_time_seconds / 3600,
    wall_hours = realtime_seconds / 3600
  )

process_summary <- task_metrics %>%
  group_by(process) %>%
  summarise(
    n_tasks = n(),
    total_wall_hours = sum(wall_hours, na.rm = TRUE),
    total_cpu_hours = sum(cpu_hours, na.rm = TRUE),
    max_peak_rss_GiB = max(peak_rss_GiB, na.rm = TRUE),
    max_peak_vmem_GiB = max(peak_vmem_GiB, na.rm = TRUE),
    total_read_GiB = sum(read_GiB, na.rm = TRUE),
    total_write_GiB = sum(write_GiB, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(total_cpu_hours >= 0.5) %>%
  arrange(desc(total_cpu_hours))

if (nrow(process_summary) == 0) {
  stop("No processes passed the 0.5 CPU-hour inclusion threshold.")
}

included_processes <- process_summary$process
task_metrics_filtered <- task_metrics %>%
  filter(process %in% included_processes)

write_tsv(
  task_metrics,
  file.path(output_dir, "Supplementary_Figure_2_task_metrics.tsv"),
  na = "NA"
)
write_tsv(
  process_summary,
  file.path(output_dir, "Supplementary_Figure_2_process_summary.tsv"),
  na = "NA"
)

process_summary
```

## Build the four-panel figure

```{r build-figure}
process_order <- process_summary %>%
  arrange(desc(total_cpu_hours)) %>%
  pull(process)

process_palette <- c(
  FastKM = "#D55E00",
  seqtk = "#E69F00",
  kmc3 = "#009E73",
  predictRF = "#0072B2",
  trainRF = "#CC79A7"
)

missing_colors <- setdiff(process_order, names(process_palette))
if (length(missing_colors) > 0) {
  stop(
    "No fixed color was defined for: ",
    paste(missing_colors, collapse = ", ")
  )
}

process_summary_plot <- process_summary %>%
  mutate(process = factor(process, levels = process_order))

task_metrics_plot <- task_metrics_filtered %>%
  mutate(process = factor(process, levels = process_order))

theme_resource <- function(base_size = 12) {
  theme_bw(base_size = base_size, base_family = "Helvetica") +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#E6E6E6", linewidth = 0.35),
      panel.border = element_rect(color = "#333333", linewidth = 0.55),
      axis.title = element_text(color = "#111111", face = "plain"),
      axis.text = element_text(color = "#3D3D3D"),
      strip.background = element_rect(fill = "#E6E6E6", color = "#333333"),
      strip.text = element_text(face = "bold", color = "#222222"),
      plot.margin = margin(7, 9, 7, 7)
    )
}

p_cpu <- process_summary_plot %>%
  mutate(process_axis = fct_reorder(process, total_cpu_hours)) %>%
  ggplot(aes(process_axis, total_cpu_hours, fill = process)) +
  geom_col(width = 0.72) +
  coord_flip() +
  scale_fill_manual(values = process_palette, drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.04))) +
  labs(x = NULL, y = "Total CPU-hours", tag = "A") +
  theme_resource()

p_memory <- process_summary_plot %>%
  mutate(process_axis = fct_reorder(process, max_peak_rss_GiB)) %>%
  ggplot(aes(process_axis, max_peak_rss_GiB, fill = process)) +
  geom_col(width = 0.72) +
  coord_flip() +
  scale_fill_manual(values = process_palette, drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.04))) +
  labs(x = NULL, y = "Maximum peak RAM (GiB)", tag = "B") +
  theme_resource()

p_scatter <- task_metrics_plot %>%
  ggplot(aes(wall_hours, peak_rss_GiB, fill = process)) +
  geom_point(
    shape = 21,
    size = 3.4,
    alpha = 0.88,
    color = "#202020",
    stroke = 0.35
  ) +
  scale_fill_manual(values = process_palette, drop = FALSE) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.05))) +
  labs(x = "Task wall time (hours)", y = "Task peak RAM (GiB)", tag = "C") +
  theme_resource()

p_io <- process_summary_plot %>%
  mutate(io_total_GiB = total_read_GiB + total_write_GiB) %>%
  select(process, io_total_GiB, total_read_GiB, total_write_GiB) %>%
  pivot_longer(
    cols = c(total_read_GiB, total_write_GiB),
    names_to = "io_type",
    values_to = "GiB"
  ) %>%
  mutate(
    io_type = recode(
      io_type,
      total_read_GiB = "Read",
      total_write_GiB = "Write"
    ),
    io_type = factor(io_type, levels = c("Read", "Write")),
    process_axis = fct_reorder(process, io_total_GiB)
  ) %>%
  ggplot(aes(process_axis, GiB, fill = process)) +
  geom_col(width = 0.7) +
  coord_flip() +
  facet_wrap(~io_type, ncol = 1, scales = "free_x") +
  scale_fill_manual(values = process_palette, drop = FALSE) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.04))) +
  labs(x = NULL, y = "Total I/O (GiB)", tag = "D") +
  theme_resource(base_size = 11)

legend_source <- p_cpu +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.key.size = unit(0.85, "lines"),
    legend.text = element_text(size = 11)
  )

shared_legend <- cowplot::get_legend(legend_source)

main_plot <- (
  (p_cpu | p_memory) /
    (p_scatter | p_io)
) +
  plot_layout(widths = c(1, 1.03), heights = c(1, 1.03)) &
  theme(
    legend.position = "none",
    plot.tag = element_text(face = "bold", size = 17, color = "#111111")
  )

combined_plot <- wrap_plots(
  main_plot,
  wrap_elements(full = shared_legend),
  ncol = 1,
  heights = c(1, 0.055)
)

combined_plot
```

## Publication-quality exports

```{r export-figure}
figure_width <- 10.5
figure_height <- 8.2

ggsave(
  filename = file.path(output_dir, "Supplementary_Figure_2.pdf"),
  plot = combined_plot,
  device = cairo_pdf,
  width = figure_width,
  height = figure_height,
  units = "in",
  bg = "white"
)

ggsave(
  filename = file.path(output_dir, "Supplementary_Figure_2_600dpi.png"),
  plot = combined_plot,
  device = ragg::agg_png,
  width = figure_width,
  height = figure_height,
  units = "in",
  dpi = 600,
  bg = "white"
)

ragg::agg_tiff(
  filename = file.path(output_dir, "Supplementary_Figure_2_600dpi.tiff"),
  width = figure_width,
  height = figure_height,
  units = "in",
  res = 600,
  compression = "lzw",
  background = "white"
)
print(combined_plot)
invisible(dev.off())
```

## Reproducibility information

```{r session-info}
sessionInfo()
```
