---
title: "Figure 1 - improved and reproducible"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
date: "2026-06-28"
---

This document reconstructs Figure 1 from local inputs. It writes the complete
figure and each individual panel to `output/`, together with the summarized
data used by the plots.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 160)

library(dplyr)
library(tidyr)
library(tibble)
library(forcats)
library(ggplot2)
library(patchwork)
library(scales)

dir.create("output/pdf", recursive = TRUE, showWarnings = FALSE)
dir.create("output/png", recursive = TRUE, showWarnings = FALSE)
dir.create("output/svg", recursive = TRUE, showWarnings = FALSE)
dir.create("output/data", recursive = TRUE, showWarnings = FALSE)

chemistry_colors <- c("R10.4" = "#F8766D", "R9.4.1" = "#00BFC4")
direction_colors <- c("Higher in R10.4" = "#F8766D",
                      "Higher in R9.4.1" = "#00BFC4")

theme_figure <- function(base_size = 10) {
  theme_minimal(base_size = base_size, base_family = "sans") +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#E6E6E6", linewidth = 0.35),
      axis.text = element_text(color = "#222222"),
      axis.title = element_text(color = "#111111"),
      strip.text = element_text(face = "bold", color = "#222222"),
      strip.background = element_rect(fill = "#F2F2F2", color = NA),
      legend.key.height = grid::unit(0.45, "cm"),
      legend.key.width = grid::unit(0.55, "cm"),
      plot.background = element_rect(fill = "white", color = NA),
      plot.margin = margin(7, 9, 7, 9)
    )
}
```

## Inputs

```{r inputs}
model_files <- c(
  HG002   = "HG002_standar_ont_sup_PAW70337.fastq_fastkm_matrix_rf_model.rds",
  HG005   = "HG005_standar_ont_sup_PAW88001.fastq_fastkm_matrix_rf_model.rds",
  HG01243 = "HG01243_1_R941_Guppy_6.1.2.fq_fastkm_matrix_rf_model.rds",
  HG01109 = "HG01109_1_R941_Guppy_6.1.2.fq_fastkm_matrix_rf_model.rds"
)

required_files <- c(model_files,
                    "Supplementary_Table_ModelPerformance.tsv",
                    "intersection_methods.rds")
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0) {
  stop("Missing input files: ", paste(missing_files, collapse = ", "))
}

models <- lapply(model_files, readRDS)

extract_gini <- function(model, dataset) {
  importance <- model$importance
  if (is.null(importance)) stop("Model ", dataset, " has no $importance matrix")
  if (!"MeanDecreaseGini" %in% colnames(importance)) {
    stop("Model ", dataset, " has no MeanDecreaseGini column")
  }
  tibble(
    feature = rownames(importance),
    dataset = dataset,
    gini = as.numeric(importance[, "MeanDecreaseGini"])
  )
}

importance_long <- bind_rows(Map(extract_gini, models, names(models))) |>
  mutate(
    dataset = factor(dataset, levels = names(model_files)),
    chemistry = if_else(dataset %in% c("HG002", "HG005"), "R10.4", "R9.4.1")
  ) |>
  group_by(dataset) |>
  mutate(gini_pct = 100 * gini / sum(gini, na.rm = TRUE)) |>
  ungroup()

write.table(importance_long, "output/data/feature_importance_by_model.tsv",
            sep = "\t", quote = FALSE, row.names = FALSE)
```

## Panel A: model performance

```{r panel-a, fig.width=6.4, fig.height=4.4}
performance <- read.delim("Supplementary_Table_ModelPerformance.tsv",
                          check.names = FALSE) |>
  mutate(
    Dataset = factor(Dataset, levels = names(model_files)),
    Chemistry = factor(Chemistry, levels = c("R10.4", "R9.4.1")),
    Class = factor(Class, levels = c("A", "B", "U"))
  )

p_a <- ggplot(performance,
              aes(x = Dataset, y = F1, color = Chemistry, shape = Class)) +
  geom_point(size = 3.0, stroke = 0.75,
             position = position_dodge(width = 0.46)) +
  scale_color_manual(values = chemistry_colors) +
  scale_shape_manual(values = c(A = 16, B = 17, U = 15)) +
  scale_y_continuous(
    limits = c(0.990, 1.0005),
    breaks = seq(0.990, 1.000, 0.0025),
    labels = label_number(accuracy = 0.001),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(x = NULL, y = "F1 score", color = "Chemistry", shape = "Class") +
  theme_figure(10) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    axis.text.x = element_text(size = 9),
    plot.margin = margin(8, 10, 4, 8)
  ) +
  guides(
    color = guide_legend(order = 1, override.aes = list(shape = 16)),
    shape = guide_legend(order = 2, override.aes = list(color = "#333333"))
  )

p_a
```

## Panel B: normalized feature importance

```{r panel-b, fig.width=7.2, fig.height=4.8}
top_n <- 15
feature_order <- importance_long |>
  group_by(feature) |>
  summarise(mean_pct = mean(gini_pct), .groups = "drop") |>
  slice_max(mean_pct, n = top_n, with_ties = FALSE) |>
  arrange(desc(mean_pct)) |>
  pull(feature)

heatmap_data <- importance_long |>
  filter(feature %in% feature_order) |>
  mutate(feature = factor(feature, levels = rev(feature_order)))

p_b <- ggplot(heatmap_data,
              aes(x = dataset, y = feature, fill = gini_pct)) +
  geom_tile(color = "white", linewidth = 0.45) +
  geom_text(aes(label = sprintf("%.1f", gini_pct)),
            color = "white", size = 2.65, fontface = "bold") +
  facet_grid(. ~ chemistry, scales = "free_x", space = "free_x") +
  scale_fill_viridis_c(option = "magma", begin = 0.08, end = 0.96,
                       name = "% total\nGini") +
  labs(x = NULL, y = NULL) +
  theme_figure(10) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 8.5),
    legend.position = "right",
    plot.margin = margin(8, 8, 4, 7)
  )

p_b
```

## Panel C: chemistry-associated shifts

```{r panel-c, fig.width=6.4, fig.height=5.2}
chemistry_delta <- importance_long |>
  group_by(chemistry, feature) |>
  summarise(mean_pct = mean(gini_pct), .groups = "drop") |>
  pivot_wider(names_from = chemistry, values_from = mean_pct) |>
  mutate(
    delta = `R10.4` - `R9.4.1`,
    abs_delta = abs(delta),
    direction = if_else(delta >= 0, "Higher in R10.4", "Higher in R9.4.1")
  ) |>
  arrange(desc(abs_delta))

write.table(chemistry_delta, "output/data/chemistry_feature_importance_delta.tsv",
            sep = "\t", quote = FALSE, row.names = FALSE)

delta_plot_data <- chemistry_delta |>
  slice_head(n = top_n) |>
  mutate(
    feature = fct_reorder(feature, delta),
    label = sprintf("%+.2f", delta),
    label_x = delta + if_else(delta >= 0, 0.08, -0.08),
    label_hjust = if_else(delta >= 0, 0, 1)
  )

delta_limit <- max(abs(delta_plot_data$delta)) + 0.55

p_c <- ggplot(delta_plot_data,
              aes(x = delta, y = feature, fill = direction)) +
  geom_col(width = 0.76) +
  geom_vline(xintercept = 0, color = "#222222", linewidth = 0.65) +
  geom_text(aes(x = label_x, label = label, hjust = label_hjust),
            size = 3.0, color = "#222222", show.legend = FALSE) +
  scale_fill_manual(values = direction_colors) +
  scale_x_continuous(
    limits = c(-delta_limit, delta_limit),
    breaks = pretty_breaks(n = 7),
    expand = expansion(mult = 0)
  ) +
  labs(x = expression(Delta*" % total Gini (R10.4 - R9.4.1)"),
       y = NULL, fill = NULL) +
  theme_figure(10) +
  theme(
    panel.grid.major.y = element_blank(),
    legend.position = "bottom",
    axis.text.y = element_text(size = 8.7),
    plot.margin = margin(7, 14, 7, 9)
  )

p_c
```

## Panel D: method agreement UpSet

The five columns are mutually exclusive read-level agreement patterns. A dark
dot means that the condition represented by that row is active in the column.

```{r panel-d, fig.width=7.2, fig.height=5.2}
method_calls <- readRDS("intersection_methods.rds")

agreement_counts <- method_calls |>
  transmute(
    pattern = case_when(
      trio_binning == hifiasm & hifiasm == rfhap ~ "All three agree",
      trio_binning == hifiasm ~ "Trio binning = Hifiasm only",
      trio_binning == rfhap ~ "Trio binning = RFHap only",
      hifiasm == rfhap ~ "Hifiasm = RFHap only",
      TRUE ~ "All different"
    )
  ) |>
  count(pattern, name = "reads") |>
  mutate(percent = 100 * reads / sum(reads))

pattern_levels <- c(
  "All three agree",
  "Trio binning = RFHap only",
  "Trio binning = Hifiasm only",
  "Hifiasm = RFHap only",
  "All different"
)

agreement_counts <- agreement_counts |>
  complete(pattern = pattern_levels, fill = list(reads = 0, percent = 0)) |>
  mutate(pattern = factor(pattern, levels = pattern_levels),
         column = as.integer(pattern)) |>
  arrange(column)

write.table(agreement_counts, "output/data/method_agreement_patterns.tsv",
            sep = "\t", quote = FALSE, row.names = FALSE)

condition_levels <- c(
  "All different",
  "All three agree",
  "Hifiasm = RFHap",
  "Trio binning = RFHap",
  "Trio binning = Hifiasm"
)

matrix_data <- crossing(
  column = seq_along(pattern_levels),
  condition = factor(condition_levels, levels = rev(condition_levels))
  ) |>
  mutate(
    active = case_when(
      column == 1 & condition %in% c("All three agree", "Hifiasm = RFHap",
                                    "Trio binning = RFHap",
                                    "Trio binning = Hifiasm") ~ TRUE,
      column == 2 & condition == "Trio binning = RFHap" ~ TRUE,
      column == 3 & condition == "Trio binning = Hifiasm" ~ TRUE,
      column == 4 & condition == "Hifiasm = RFHap" ~ TRUE,
      column == 5 & condition == "All different" ~ TRUE,
      TRUE ~ FALSE
    ),
    row = as.integer(condition)
  )

segments <- matrix_data |>
  filter(active) |>
  group_by(column) |>
  summarise(ymin = min(row), ymax = max(row), .groups = "drop") |>
  filter(ymin != ymax)

bar_max <- max(agreement_counts$reads)

d_bar <- ggplot(agreement_counts, aes(x = column, y = reads)) +
  geom_col(width = 0.62, fill = "#3B73B9") +
  geom_text(aes(label = paste0(comma(reads), "\n", sprintf("%.1f%%", percent))),
            vjust = -0.18, size = 2.8, lineheight = 0.9) +
  scale_x_continuous(breaks = seq_along(pattern_levels), labels = NULL,
                     limits = c(0.5, length(pattern_levels) + 0.5),
                     expand = expansion(mult = 0)) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()),
                     limits = c(0, bar_max * 1.17),
                     expand = expansion(mult = c(0, 0))) +
  labs(x = NULL, y = "Reads") +
  theme_figure(9.5) +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.margin = margin(6, 8, 0, 6)
  )

d_matrix <- ggplot(matrix_data, aes(x = column, y = condition)) +
  geom_tile(aes(fill = as.integer(condition) %% 2),
            width = 1, height = 1, alpha = 0.05, show.legend = FALSE) +
  geom_segment(data = segments,
               aes(x = column, xend = column, y = ymin, yend = ymax),
               inherit.aes = FALSE, color = "#3A3A3A", linewidth = 0.8) +
  geom_point(aes(color = active), size = 2.7) +
  scale_color_manual(values = c(`TRUE` = "#3A3A3A", `FALSE` = "#D9D9D9"),
                     guide = "none") +
  scale_fill_gradient(low = "white", high = "#999999", guide = "none") +
  scale_x_continuous(
    breaks = seq_along(pattern_levels),
    labels = c("All agree", "Trio-RFHap", "Trio-Hifiasm",
               "Hifiasm-RFHap", "All different"),
    limits = c(0.5, length(pattern_levels) + 0.5),
    expand = expansion(mult = 0)
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 9.5) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_text(color = "#222222", size = 8.5),
    axis.text.x = element_text(color = "#222222", size = 7.7,
                               angle = 28, hjust = 1),
    axis.ticks = element_blank(),
    plot.margin = margin(0, 8, 9, 6)
  )

p_d_inner <- d_bar / d_matrix + plot_layout(heights = c(0.63, 0.37))
p_d <- wrap_elements(full = p_d_inner)
p_d
```

## Complete figure

```{r complete-figure, fig.width=13.5, fig.height=10.5}
figure_1 <- ((p_a | p_b) / (p_c | p_d)) +
  plot_layout(widths = c(0.95, 1.15), heights = c(0.92, 1.08)) +
  plot_annotation(
    tag_levels = "A",
    theme = theme(
      plot.tag = element_text(size = 16, face = "bold", color = "#111111"),
      plot.tag.position = c(0.006, 0.995),
      plot.background = element_rect(fill = "white", color = NA),
      plot.margin = margin(5, 5, 5, 5)
    )
  )

figure_1

ggsave("output/pdf/Figure1_improved.pdf", figure_1,
       width = 13.5, height = 10.5, units = "in", device = cairo_pdf,
       bg = "white")
ggsave("output/png/Figure1_improved.png", figure_1,
       width = 13.5, height = 10.5, units = "in", dpi = 300,
       bg = "white")
ggsave("output/svg/Figure1_improved.svg", figure_1,
       width = 13.5, height = 10.5, units = "in", device = svg,
       bg = "white")

panels <- list(A = p_a, B = p_b, C = p_c, D = p_d)
panel_sizes <- list(A = c(6.4, 4.4), B = c(7.2, 4.8),
                    C = c(6.4, 5.2), D = c(7.2, 5.2))
for (panel_name in names(panels)) {
  dims <- panel_sizes[[panel_name]]
  ggsave(file.path("output/pdf", paste0("Figure1", panel_name, ".pdf")),
         panels[[panel_name]], width = dims[1], height = dims[2],
         units = "in", device = cairo_pdf, bg = "white")
  ggsave(file.path("output/png", paste0("Figure1", panel_name, ".png")),
         panels[[panel_name]], width = dims[1], height = dims[2],
         units = "in", dpi = 300, bg = "white")
}
```

## Session information

```{r session-info}
sessionInfo()
```
