---
title: "Supplementary Figure 4"
subtitle: "RFhap model performance across datasets"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
date: "2026-07-03"
---

## 1. Load packages and create output directories

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 160
)

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

dir.create("output/pdf", recursive = TRUE, showWarnings = FALSE)
dir.create("output/png", recursive = TRUE, showWarnings = FALSE)
```

## 2. Read model-performance data

The input table must be located in the same directory as this R Markdown
document. Sensitivity is not included because it is mathematically identical
to recall (`TP / (TP + FN)`).

```{r read-data}
metrics_df <- read.csv(
  "Supplementary_Table_ModelPerformance.csv",
  check.names = TRUE
)

head(metrics_df)
```

## 3. Select and reshape the metrics

The five metrics are ordered according to their final position in the figure:
F1 score, balanced accuracy, and precision in the upper row; recall and
specificity in the lower row.

```{r reshape-data}
metric_order <- c(
  "F1",
  "Balanced Accuracy",
  "Precision",
  "Recall",
  "Specificity"
)

plot_df <- metrics_df |>
  select(
    Dataset,
    Chemistry,
    Class,
    F1,
    Balanced.Accuracy,
    Precision,
    Recall,
    Specificity
  ) |>
  rename(`Balanced Accuracy` = Balanced.Accuracy) |>
  pivot_longer(
    cols = all_of(metric_order),
    names_to = "Metric",
    values_to = "Value"
  ) |>
  mutate(
    Metric = factor(Metric, levels = metric_order)
  )

head(plot_df)
```

## 4. Define a function for the individual panels

All panels use the same scale, colors, shapes, and graphical theme. Dataset
labels are displayed only in the lower row to reduce repetition.

```{r panel-function}
make_metric_plot <- function(metric_name, show_x = FALSE) {

  ggplot(
    filter(plot_df, Metric == metric_name),
    aes(
      x = Dataset,
      y = Value,
      shape = Class,
      color = Chemistry
    )
  ) +
    geom_point(
      size = 3,
      position = position_dodge(width = 0.45)
    ) +
    scale_y_continuous(
      breaks = seq(0.98, 1.00, by = 0.005),
      labels = scales::label_number(accuracy = 0.001)
    ) +
    coord_cartesian(ylim = c(0.98, 1.00)) +
    labs(
      title = metric_name,
      x = NULL,
      y = NULL,
      color = "Chemistry",
      shape = "Class"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = if (show_x) {
        element_text(angle = 30, hjust = 1)
      } else {
        element_blank()
      },
      axis.ticks.x = if (show_x) {
        element_line()
      } else {
        element_blank()
      },
      plot.title = element_text(
        face = "bold",
        hjust = 0.5,
        size = 12
      ),
      plot.margin = margin(6, 7, 6, 7)
    )
}
```

## 5. Build the five metric panels

```{r build-panels}
p_f1 <- make_metric_plot("F1")
p_balanced <- make_metric_plot("Balanced Accuracy")
p_precision <- make_metric_plot("Precision")
p_recall <- make_metric_plot("Recall", show_x = TRUE)
p_specificity <- make_metric_plot("Specificity", show_x = TRUE)
```

## 6. Arrange the panels

The layout uses twelve equal columns. Each panel occupies four columns. The
lower row is shifted slightly to the left to achieve better optical balance.

```{r arrange-panels}
figure_design <- "
AAAABBBBCCCC
#DDDDEEEE###
"

plots_grid <-
  p_f1 +
  p_balanced +
  p_precision +
  p_recall +
  p_specificity +
  plot_layout(
    design = figure_design,
    guides = "collect",
    heights = c(1, 1)
  ) &
  theme(legend.position = "bottom")

shared_y_label <- wrap_elements(
  full = grid::textGrob(
    "Metric value",
    rot = 90,
    gp = grid::gpar(fontsize = 12)
  )
)

p_perf <-
  shared_y_label +
  plots_grid +
  plot_layout(widths = c(0.035, 1)) +
  plot_annotation(
    title = "RFhap model performance across datasets",
    theme = theme(
      plot.title = element_text(
        face = "bold",
        hjust = 0.5,
        size = 16
      )
    )
  )
```

## 7. Display Supplementary Figure 4

```{r supplementary-figure-4, echo=FALSE, fig.width=9, fig.height=6.8, out.width="100%"}
p_perf
```

## 8. Export PDF and PNG versions

```{r export-figure}
ggsave(
  filename = "output/pdf/Supplementary_Figure_4.pdf",
  plot = p_perf,
  width = 9,
  height = 6.8,
  units = "in",
  device = cairo_pdf,
  bg = "white"
)

ggsave(
  filename = "output/png/Supplementary_Figure_4.png",
  plot = p_perf,
  width = 9,
  height = 6.8,
  units = "in",
  dpi = 400,
  bg = "white"
)
```

## 9. Session information

```{r session-info}
sessionInfo()
```
