---
title: "Figure 3 - chromosomal distribution of long-switch errors"
output:
  html_document:
    toc: true
    code_folding: show
date: "2026-06-28"
---

This document composes the maternal and paternal chromosome maps as a mirrored
standalone main figure. The source maps are local PNG files generated with
`karyoploteR`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 160)
library(png)
library(grid)
library(ggplot2)
library(patchwork)

dir.create("output/pdf", recursive = TRUE, showWarnings = FALSE)
dir.create("output/png", recursive = TRUE, showWarnings = FALSE)
dir.create("output/svg", recursive = TRUE, showWarnings = FALSE)
```

## Prepare mirrored chromosome tracks

The rasterized chromosome labels and the original small legend are removed.
Only the maternal chromosome tracks are horizontally mirrored, while chromosome
labels and the shared legend are added as vector text and symbols.

```{r load-images}
prepare_track_map <- function(path, mirror = FALSE, remove_legend = FALSE) {
  img <- readPNG(path)
  if (length(dim(img)) != 3 || dim(img)[3] < 3) {
    stop("Expected an RGB or RGBA PNG: ", path)
  }

  rgb <- img[, , 1:3, drop = FALSE]
  if (dim(img)[3] == 4) {
    alpha <- img[, , 4]
    for (channel in seq_len(3)) {
      rgb[, , channel] <- rgb[, , channel] * alpha + (1 - alpha)
    }
  }

  # Remove the original legend from the paternal raster; the replacement below
  # is vector-based and shared by both haplotypes.
  if (remove_legend) rgb[1450:1710, 920:1260, ] <- 1

  # Common crop preserves identical vertical coordinates in both maps.
  rgb <- rgb[55:1965, 145:1550, , drop = FALSE]
  if (mirror) rgb <- rgb[, dim(rgb)[2]:1, , drop = FALSE]
  rgb
}

maternal_map <- prepare_track_map("hapA_karyoplot.png", mirror = TRUE)
paternal_map <- prepare_track_map("hapB_karyoplot.png", remove_legend = TRUE)

# Track centers in the original 2,000 px karyoploteR canvas. The spacing is
# fixed by the shared plot parameters used to generate both source maps.
chromosome_centers <- seq(89.75, 1919.75, length.out = 22)
chromosome_labels <- data.frame(
  chromosome = paste0("chr", 1:22),
  y_source = 1 - (chromosome_centers - 55) / (1965 - 55)
) |>
  transform(y = 0.075 + 0.875 * y_source)
```

## Compose Figure 3

```{r figure3, fig.width=13.5, fig.height=8.2}
legend_data <- data.frame(
  label = c("RFhap", "Hifiasm", "Telomeres", "Centromeres"),
  color = c("#FC8D62", "#66C2A5", "#0000FF", "#FFD92F"),
  x = c(0.30, 0.42, 0.55, 0.68)
)

figure_3 <- ggplot() +
  annotation_custom(
    rasterGrob(maternal_map, interpolate = TRUE,
               width = grid::unit(1, "npc"), height = grid::unit(1, "npc")),
    xmin = 0.015, xmax = 0.495, ymin = 0.075, ymax = 0.95
  ) +
  annotation_custom(
    rasterGrob(paternal_map, interpolate = TRUE,
               width = grid::unit(1, "npc"), height = grid::unit(1, "npc")),
    xmin = 0.505, xmax = 0.985, ymin = 0.075, ymax = 0.95
  ) +
  geom_text(
    data = chromosome_labels,
    aes(x = 0.5, y = y, label = chromosome),
    size = 2.8, color = "#222222"
  ) +
  annotate("text", x = 0.245, y = 0.985, label = "Maternal haplotype",
           fontface = "bold", size = 4.1) +
  annotate("text", x = 0.755, y = 0.985, label = "Paternal haplotype",
           fontface = "bold", size = 4.1) +
  geom_segment(
    data = legend_data,
    aes(x = x, xend = x + 0.018, y = 0.025, yend = 0.025, color = color),
    linewidth = 3.2, lineend = "butt", show.legend = FALSE
  ) +
  geom_text(
    data = legend_data,
    aes(x = x + 0.024, y = 0.025, label = label),
    hjust = 0, size = 3.4, color = "#222222"
  ) +
  scale_color_identity() +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE,
                  clip = "off") +
  theme_void(base_size = 11) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(6, 8, 6, 8)
  )

figure_3

ggsave("output/pdf/Figure3_chromomap.pdf", figure_3,
       width = 13.5, height = 8.2, units = "in", device = cairo_pdf,
       bg = "white")
ggsave("output/png/Figure3_chromomap.png", figure_3,
       width = 13.5, height = 8.2, units = "in", dpi = 300,
       bg = "white")
ggsave("output/svg/Figure3_chromomap.svg", figure_3,
       width = 13.5, height = 8.2, units = "in", device = svg,
       bg = "white")
```

```{r session-info}
sessionInfo()
```
